Prevailing scenarios of functional change in Anthropocene bird and mammal communities

### 2023-06-23 paper/GEB Submission/manuscript_GEB.tex
Final submission tweaks.

Methods: We paired thousands of bird and mammal assemblage time series
from the BioTIME database with existing trait data representative of a
species' functional role to reconstruct time series of functional
diversity metrics. Using generalized linear mixed models we estimated
general trends in those metrics and trends for individual studies.

Results: We found no overall trend in any functional diversity metric,
despite data replicating species-based patterns of constant richness
with increasing turnover. The lack of trend held even after correcting
for changes in species richness. At the study-level, we identified four
prevailing scenarios of species and functional change, which showed
links to the duration of the observation window.

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevelant than the other across many
taxa, biomes, and realms. At the study-level, there were also a
substantial number of time series exhibiting no species or functional
change, however the majority of studies showed a shift in a species or
functional metric. With no one prevailing scenario of change, it will be
critical to link change scenarios to drivers of change, particularly to
identify communities with capacity to resist drivers from those not
experiencing substantial pressure from a driver.
\end{abstract}

@@

The meaning of general trends derived from limited data, and their
relevance for conservation, is a topic of on going debate. Global
analyses have been heavily criticized for geographic biases, lack of
data in the most heavily impacted areas, and not taking into account the
ecological context of individual studies (10--12). Many of these
criticisms reflect limitations of ecological data on the whole, and
proponents argue we must make the best of the imperfect data available
(13). Still, there is general consensus on a few local-scale patterns
that seem to be characteristic of the anthropocene, including the
prevalence of species turnover across communities and patterns of both
richness increase and decrease in different taxa and biomes (14).
Understanding the implications of change in species assemblage for
ecosystem stability and function is a critical next step in moving
forward the biodiversity change conversation.

@@

Here we perform the first multi-taxa, multi-realm assessment of
functional diversity change through time. We focus on mammal and bird
species as a significant subset of the world's biodiversity heavily
impacted by anthropogenic change. While examining trends in plants,
invertebrates, and other vertebrate species is of equal interest, trait
data for those taxa raise additional challenges such as limited and
biased species coverage (45), a lack of accepted species-level means,
and differences in the types of traits collected. In order to ensure
comparability across taxa in trait type and data quality we therefore
focus on mammals and birds. Traits were intentionally selected to be
representative of a species' Eltonian niche, thereby summarizing the
functional role they play in the community (46). An initial assessment
of amphibian trends is included in the supplement, but excluded from
general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not there is a general trend of functional
change, both in observed metrics and in metrics corrected for changes in
species richness. We further assess the prevailing modes of change
occurring at the study level to determine how functional change is
related to species patterns, and which kinds of species and functional
change are most common across communities. We distinguish between three
possible scenarios of functional change: 1) significant functional
change accompanied by changes in species-based metrics, 2) maintenance
of functional structure despite significant species changes, 3)
maintenance of functional and species metrics. Based on expectation due
to human impacts, we expect to find significant functional changes,
often accompanied by species change.

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high quality assemblage time series. All studies
included in the database follow consistent sampling protocols and
represent full assemblages rather than populations of single species
(36). All time series include abundance of observed species. Following
best practices for the database (47), studies with multiple sample
locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 36 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

@@

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (52, 53). The number of trait axes was limited to the maximum
number of traits that fulfills the criteria \(s >= 2^t\), where \(s\) is
the number of species and \emph{t} is the number of traits. This
restriction allows for a sufficient number of axes to capture the trait
space while maintaining computational feasibility (54). Metrics
incorporated weighting based on species abundance.

@@

To test for trends within and between different levels of taxa, biome,
and realm we fit separate models with each of those factors added as a
predictor to the original model structure. We estimated within-level
slopes and calculated between-level contrasts using the \emph{emmeans}
package (v1.8.2, 56). We assessed the impact of time series duration and
start year on study-level trends using linear models with duration and
start year as predictors. We also assessed the impact of potential
study-level outliers on general trends. Details on that process can be
found in the supplement. All models in our analysis were fit using the
\emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
by Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(57--59).

@@

We did find significant differences between taxa, realms, and climate
for Jaccard similarity and some of the \emph{CWM's}. For example while
Jaccard similarity was decreasing in the general trend and there were
significant within group slopes for Mammals, Birds, and Terrestrial
communities, there was no significant slope for the marine realm,
indicating that the general trend is mostly driven by turnover in
terrestrial communities. For bird communities, we also found within
group trends and between group differences for trends in foraging
behavior. We found a significant increasing trend in utilization of the
canopy in Tropical communities that was significantly different from the
trends for Polar/Temperate and Temperate communities. There was a
significant decrease in utilization of the understory for Terrestrial
communities and significant increase in foraging below the water surface
for global studies (but see the previous limitations of Global data).
There was also a significant positive slope for \emph{CWM} body mass for
the single Temperate/Tropical study (with three time series) which were
marine mammal communities.

@@

At the study level, biodiversity change fell into four dominant
scenarios of change based on trends in richness, turnover, and
functional metrics: no change, only functional change, richness and
functional change with turnover, and richness and functional change
without turnover (Table \ref{tab:changeScenarios}). Around a quarter of
studies experienced no trend in richness, turnover, or functional
metrics. These studies spanned the distribution of study durations
excluding only the very longest running studies, with the longest no
change time series lasting 23 years.

Another quarter of the studies showed no trend in richness or turnover,
but did show a significant shift in at least one summary functional
metric. These communities are best aligned with scenarios of species
replacement where only a few functional outliers replaced functionally
indistinct species or vice versa (there were studies with positive and
negative trends in all three functional metrics). Studies in the
scenario of only functional change were heavily skewed toward shorter
running time series, indicating that they may represent a limited
snapshot of communities that on longer time scales would be exhibiting
significant change in species based metrics, particularly as species
replacement continued adding to turnover. Communities in this scenario,
and the previous scenario of no functional change are consistent with
hypotheses that species richness may be strongly regulated through time
(3, 8), and emphasize that maintenance of functional structure can be
divorced from those processes.

@@

Our results are consistent with ecological expectations that community
dissimilarity would be greater the larger the observation time window
purely due to background processes, which is likely further accelerated
by global change drivers (14). While we do not address how observed
rates differ from background expectation here, we can make some
assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a
directional change in the functional space. Species are not therefore
being replaced by exact functional counter parts. The fact that we
observed both increases and decreases in corrected functional richness
accompanying turnover is a reminder that replacements are not always an
indication of functional loss, but can also lead to increased functional
richness. While shifts in functional space may be the result of
environmental pressure, we cannot distinguish between directional change
and shifts due to the unavailability of an analog species in the species
pool.

@@

Still, some results stand in contrast to predictions for trait shifts
under global change. For example, mean body size is predicted to
decrease as a result of climate change impacts and megafaunal loss (67),
a phenomena which has already been well documented empirically and
experimentally in multiple taxa (68--71). While the species-level trait
means used here are not appropriate for assessing intraspecific body
size shifts, we would expect to see shifts in community-wide means due
to local losses of large-bodied species. Instead, we found no evidence
of a trend in \emph{CWM} body size, with the exception of a single study
of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have
experienced significant loss of large-bodied species before the
observation window, with contemporary loss rates slowing (72). Trends
could be significantly different for the same time periods in regions of
sub-Saharan Africa, for example, which has poor representation in our
dataset but where megafauna exist on the landscape and are increasingly
threatened (73).

@@

We present here four prevailing scenarios of change experienced in bird
and mammal communities. While we offer an first discussion of which
kinds of change are most common and why, assessing the true global
prevalence will require continued efforts to fill data gaps, a well
recognized challenge in ecology (6, 12, 77). Still, a ``scenarios of
change'' framework can provide structure for future work addressing
functional shifts, particularly as we reconcile results from broad-scale
syntheses with in-depth single system studies. It will be particularly
critical to link forms of change to individual drivers to assess which
drivers may impact species and functional diversity differently.
Understanding those links will help identify where directly measuring
functional structure instead of just species change is necessary for
understand impacts on a given system. In addition, direct links to
drivers will improve our ability to distinguish between communities
experiencing no change due to a lack of perturbation from communities
with high resilience in the face of disturbance.

>>>>

Methods: We paired thousands of bird and mammal assemblage time series
from the BioTIME database with existing trait data representative of a
species' functional role to reconstruct time series of functional
diversity metrics. Using generalized linear mixed models, we estimated
general trends in those metrics and trends for individual studies.

Results: We found no overall trend in any functional diversity metric,
despite data replicating species-based patterns of constant richness
with increasing turnover. The lack of trend held even after correcting
for changes in species richness. At the study-level, there were also a
substantial number of time series exhibiting no species or functional
change, however most studies showed a shift in a species or functional
metric.

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. At the study-level, we identified four
prevailing scenarios of species and functional change, which showed
links to the duration of the observation window. With no one prevailing
scenario of change, it will be critical to link change scenarios to
drivers of change, particularly to identify communities with capacity to
resist drivers from those not experiencing substantial pressure from a
driver.
\end{abstract}

@@

The meaning of general trends derived from limited data, and their
relevance for conservation, is a topic of on going debate. Global
analyses have been heavily criticized for geographic biases, lack of
data in the most heavily impacted areas, and not considering the
ecological context of individual studies (10--12). Many of these
criticisms reflect limitations of ecological data on the whole, and
proponents argue we must make the best of the imperfect data available
(13). Still, there is general consensus on a few local-scale patterns
that seem to be characteristic of the anthropocene, including the
prevalence of species turnover across communities and patterns of both
richness increase and decrease in different taxa and biomes (14).
Understanding the implications of change in species assemblage for
ecosystem stability and function is a critical next step in moving
forward the biodiversity change conversation.

@@

Here we perform the first multi-taxa, multi-realm assessment of
functional diversity change through time. We focus on mammal and bird
species as a significant subset of the world's biodiversity heavily
impacted by anthropogenic change. While examining trends in plants,
invertebrates, and other vertebrate species is of equal interest, trait
data for those taxa raise additional challenges such as limited and
biased species coverage (45), a lack of accepted species-level means,
and differences in the types of traits collected. To ensure
comparability across taxa in trait type and data quality we therefore
focus on mammals and birds. Traits were intentionally selected to be
representative of a species' Eltonian niche, thereby summarizing the
functional role they play in the community (46). An initial assessment
of amphibian trends is included in the supplement, but excluded from
general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not there is a general trend of functional
change, both in observed metrics and in metrics corrected for changes in
species richness. We further assess the prevailing modes of change
occurring at the study level to determine how functional change is
related to species patterns, and which kinds of species and functional
change are most common across communities. We expect to find significant
functional changes happening alongside and independent of species
richness changes as a result of ongoing anthropogenic impacts and global
change drivers.

@@

We obtained mammal and bird time series from the BioTIME database, a
global repository of high-quality assemblage time series. All studies
included in the database follow consistent sampling protocols and
represent full assemblages rather than populations of single species
(36). All time series include abundance of observed species. Following
best practices for the database (47), studies with multiple sample
locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 36 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

@@

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (52, 53). The number of trait axes was limited to the maximum
number of traits that fulfills the criteria \(s >= 2^t\), where \(s\) is
the number of species and \emph{t} is the number of traits. This
restriction allows for enough axes to capture the trait space while
maintaining computational feasibility (54). Metrics incorporated
weighting based on species abundance.

@@

To test for trends within and between different levels of taxa, biome,
and realm we fit separate models with each of those factors added as a
predictor to the original model structure. We estimated within-level
slopes and calculated between-level contrasts using the \emph{emmeans}
package (v1.8.2, 56). We assessed the impact of time series duration and
start year on study-level trends using linear models with duration and
start year as predictors. All models in our analysis were fit using the
\emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
by Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(57--59).

@@

We did find significant differences between taxa, realms, and climate
for Jaccard similarity and some of the \emph{CWM's}. For example, while
Jaccard similarity was decreasing in the general trend and there were
significant within group slopes for Mammals, Birds, and Terrestrial
communities, there was no significant slope for the marine realm,
indicating that the general trend is mostly driven by turnover in
terrestrial communities. For bird communities, we also found within
group trends and between group differences for trends in foraging
behavior. We found a significant increasing trend in utilization of the
canopy in Tropical communities that was significantly different from the
trends for Polar/Temperate and Temperate communities. There was a
significant decrease in utilization of the understory for Terrestrial
communities and significant increase in foraging below the water surface
for global studies (but see the previous limitations of Global data).
There was also a significant positive slope for \emph{CWM} body mass for
the single Temperate/Tropical study (with three time series) which were
marine mammal communities.

@@

At the study level, biodiversity change fell into four dominant
scenarios of change based on trends in richness, turnover, and
functional metrics: no change, only functional change, richness and
functional change with turnover, and richness and functional change
without turnover (Table \ref{tab:changeScenarios}). In contrast to our
expectation based on anthropogenic impacts, around a quarter of studies
experienced no trend in richness, turnover, or functional metrics. These
studies spanned the distribution of study durations excluding only the
very longest running studies, with the longest no change time series
lasting 23 years.

Another quarter of the studies showed no trend in richness or turnover,
but did show a significant shift in at least one summary functional
metric. These communities are best aligned with scenarios of species
replacement where only a few functional outliers replaced functionally
indistinct species or vice versa (there were studies with positive and
negative trends in all three functional metrics). Studies in the
scenario of only functional change were heavily skewed toward shorter
running time series, indicating that they may represent a limited
snapshot of communities that on longer time scales would be exhibiting
significant change in species-based metrics, particularly as species
replacement continued adding to turnover. Communities in this scenario,
and the previous scenario of no functional change are consistent with
hypotheses that species richness may be strongly regulated through time
(3, 8), and emphasize that maintenance of functional structure can be
divorced from those processes.

@@

Our results are consistent with ecological expectations that community
dissimilarity would be greater the larger the observation time window
purely due to background processes, which is likely further accelerated
by global change drivers (14). While we do not address how observed
rates differ from background expectation here, we can make some
assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a
directional change in the functional space. Species are not therefore
being replaced by exact functional counter parts. The fact that we
observed both increases and decreases in corrected functional richness
accompanying turnover is a reminder that replacements are not always an
indication of functional loss but can also lead to increased functional
richness. While shifts in functional space may be the result of
environmental pressure, we cannot distinguish between directional change
and shifts due to the inavailability of an analog species in the species
pool.

@@

Still, some results stand in contrast to predictions for trait shifts
under global change. For example, mean body size is predicted to
decrease as a result of climate change impacts and megafaunal loss (67),
a phenomenon which has already been well documented empirically and
experimentally in multiple taxa (68--71). While the species-level trait
means used here are not appropriate for assessing intraspecific body
size shifts, we would expect to see shifts in community-wide means due
to local losses of large-bodied species. Instead, we found no evidence
of a trend in \emph{CWM} body size, with the exception of a single study
of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have
experienced significant loss of large-bodied species before the
observation window, with contemporary loss rates slowing (72). Trends
could be significantly different for the same time periods in regions of
sub-Saharan Africa, for example, which has poor representation in our
dataset but where megafauna exist on the landscape and are increasingly
threatened (73).

@@

We present here four prevailing scenarios of change experienced in bird
and mammal communities. While we offer a first discussion of which kinds
of change are most common and why, assessing the true global prevalence
will require continued efforts to fill data gaps, a well recognized
challenge in ecology (6, 12, 77). Still, a ``scenarios of change''
framework can provide structure for future work addressing functional
shifts, particularly as we reconcile results from broad-scale syntheses
with in-depth single system studies. It will be particularly critical to
link forms of change to individual drivers to assess which drivers may
impact species and functional diversity differently. Understanding those
links will help identify where directly measuring functional structure
instead of just species change is necessary for understand impacts on a
given system. In addition, direct links to drivers will improve our
ability to distinguish between communities experiencing no change due to
a lack of perturbation from communities with high resilience in the face
of disturbance.
